Using Monte Carlo Algorithms to Achieve Resiliency \& Performance at Scale for Linear and Nonlinear Solver Applications

### 2016-01-26 doc/ecp/pres.tex
Added slide content.

  \item Models defined by PDEs that result in sparse linear systems
    \begin{itemize}
    \item fluid flow
    \item radiation transport
    \item electrostatics
    \end{itemize}

@@

@@

\begin{frame}{Limitations of Monte Carlo}
  \begin{itemize}
  \item Solving linear systems of equations with ``pure'' Monte Carlo methods
    is generally intractable
    \begin{itemize}
    \item Central limit theorem is barrier to accurate solutions
    \end{itemize}
    \begin{figure}
      \centering
      \includegraphics[width=1.75in]{../NCState_seminar_2014/Ifiss_ConvDiff}
    \end{figure}
  \item There is a potentially restrictive limit on the spectral properties of
    $\mathbf{H}$
    \begin{equation}
      \rho(\mathbf{H}) < 1\:, \rho(\hat{\mathbf{H}}) < 1\:, \hat{H}_{ij} =
      \frac{H_{ij}^2}{P_{ij}}
    \end{equation}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Preconditioning}
  \begin{itemize}
    \item Preconditioning for Monte Carlo introduces nonstandard considerations
      \begin{itemize}
        \item Elements of preconditioned matrix must be explicitly available
        \item Goal of preconditioning is to reduce $\rho(\hat{\mathbf{H}})$,
          may not follow ``standard'' trends
      \end{itemize}
  \end{itemize}
  \vfill

@@

  \begin{table}[!h]
  \centering
  \begin{tabular}{|c|c|c|c|}
  \hline
  \textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
  &$\rho(\hat{H})$\\
  \hline
   0.05 & 8.2797 & $0.9610$ & $1.0231$\\
  \hline
   0.01 & 33.1390 & $0.8668$ & $0.8279$\\
  \hline
  \end{tabular}
  \caption{$\rho(H)$ and $\rho(\hat{H})$ for 2D
  Laplacian with AINV preconditioning.}
  \label{tab:lap2d_ainv}
  \end{table}

@@

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Each parallel process owns a piece of the domain (linear
        system)
        \bigskip
      \item Random walks must be transported between adjacent domains
        through parallel communication
        \bigskip
      \item Domain decomposition determined by the input system
        \bigskip
      \item Load balancing not yet addressed
      \end{itemize}
    \end{column}

@@

\begin{frame}

@@

  {\bf MCLS} (\underline{github.com/sslattery/MCLS})
  \begin{itemize}
  \item Massively parallel (MPI) implementation of domain decomposed MCSA
    algorithm
  \item Implemented in Trilinos framework and callable through Trilinos'
    abstract numerical algorithms (Thyra, Stratimikos) interface
  \item Works with any Tpetra row matrix
  \end{itemize}
  \vfill
  {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
  \begin{itemize}
  \item GPU-accelerated on-node MC solve
  \item Domain-local MC solves implemented for distributed systems as an
    additive Schwartz method
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\section{Transition and Roadmap}
%%---------------------------------------------------------------------------%%

@@

>>>>

  \item Models defined by PDEs that result in sparse linear systems, examples:
    \begin{itemize}
    \item fluid flow
    \item radiation transport
    \item gas dynamics
    \end{itemize}

@@

\begin{frame}{Summary}

  Much of the first 2 years of the project were attempting to resolve
  \textit{algorithmic} challenges related to solver performance and efficiency
  \begin{itemize}
  \item Iterative schemes to improve efficiency
  \item Analysis to determine convergence
  \item Development and application of effective preconditioners
  \end{itemize}

  We have also developed scalable parallel algorithms that can take advantage
  of the benefits of replicating Monte Carlo random walks
  \begin{itemize}
  \item Developed multilevel parallel strategies for domain decomposed Monte
    Carlo that allows redundancy
  \item Developed concurrent on-node random walk algorithms for GPU
    optimization
  \end{itemize}

  We have developed two open source software packages that contain
  production-quality versions of these solvers (MPI and GPU).

@@

\begin{frame}{Limitations of Monte Carlo}
  \begin{itemize}
  \item Solving linear systems of equations with ``pure'' Monte Carlo methods
    is generally intractable
    \begin{itemize}
    \item Central limit theorem is barrier to accurate solutions
    \end{itemize}
    \begin{figure}
      \centering
      \includegraphics[width=1.75in]{../NCState_seminar_2014/Ifiss_ConvDiff}
    \end{figure}
  \item There is a potentially restrictive limit on the spectral properties of
    $\mathbf{H}$
    \begin{equation*}
      \rho(\mathbf{H}) < 1\:, \rho(\hat{\mathbf{H}}) < 1\:, \hat{H}_{ij} =
      \frac{H_{ij}^2}{P_{ij}}
    \end{equation*}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Preconditioning}
  \begin{itemize}
    \item Preconditioning needed to address convergence constraints due to
      spectral radius conditions
    \item Preconditioning for Monte Carlo introduces nonstandard considerations
      \begin{itemize}
      \item Elements of preconditioned matrix must be explicitly available,
        sparsity in $\mathbf{A}$, $\mathbf{M}$, does not imply sparsity in
        $\mathbf{M^{-1}A}$
        \item Goal of preconditioning is to reduce $\rho(\hat{\mathbf{H}})$,
          may not follow ``standard'' trends
      \end{itemize}
  \end{itemize}
  \vfill

@@

  \begin{table}[!h]
  \centering
  \begin{tabular}{|c|c|c|c|}
  \hline
  \textbf{Drop tolerance} & \textbf{$\frac{nnz(H)}{nnz(A)}$}& $\rho(H)$
  &$\rho(\hat{H})$\\
  \hline
   0.05 & 8.2797 & $0.9610$ & $1.0231$\\
  \hline
   0.01 & 33.1390 & $0.8668$ & $0.8279$\\
  \hline
  \end{tabular}
  \caption{$\rho(H)$ and $\rho(\hat{H})$ for 2D
  Laplacian with AINV preconditioning.}
  \label{tab:lap2d_ainv}
  \end{table}
\end{frame}

@@

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Each parallel process owns a piece of the domain (linear
        system)
        \bigskip
      \item Random walks must be transported between adjacent domains
        through parallel communication
        \bigskip
      \item Domain decomposition determined by the input system
        \bigskip
      \item Load balancing not yet addressed
        \bigskip
      \item Uses modified version of the Milagro algorithm (LANL)
      \end{itemize}
    \end{column}

@@

\begin{frame}{Remaining Research}

@@

  {\bf MCLS} (\underline{github.com/sslattery/MCLS})
  \begin{itemize}
  \item Massively parallel (MPI) implementation of domain decomposed MCSA
    algorithm
  \item Implemented in Trilinos framework and callable through Trilinos'
    abstract numerical algorithms (Thyra, Stratimikos) interface
  \item Works with any Tpetra row matrix
  \end{itemize}
  \vfill
  {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
  \begin{itemize}
  \item GPU-accelerated on-node MC solve
  \item Domain-local MC solves implemented for distributed systems as an
    additive Schwartz method
  \end{itemize}
  \vfill
  All software is BSD open source.
  \vfill
  Uses TriBITS (Trilinos) open source build system
  \vfill
  Both software packages contain a complete suite of unit-tests integrated
  through GoogleTest (Alea) or TeuchosTest (MCLS)
\end{frame}

\begin{frame}{Publications and Presentations}

  \underline{Publications}
  \begin{itemize}
  \item M. Benzi, T.M. Evans, S.P. Hamilton, M.L. Pasini, and
    S.R. Slattery. Analysis of Monte Carlo Accelerated Iterative Methods for
    Sparse Linear Systems. \textit{in preparation}, 2016.
  \item S.R. Slattery, T.M. Evans, P.P.H. Wilson. A Spectral Analysis of the
    Domain Decomposed Monte Carlo Method for Linear Systems. \textit{Nuclear
      Engineering Design},  {\bf 295}, 632--638, 2015.
  \item T.M. Evans, S.W. Mosher, S.R. Slattery, and S.P. Hamilton. A Monte Carlo
    Synthetic Acceleration Method for Solving the Thermal Radiation Diffusion
    Equation. \textit{J. Comp. Phys.}, {\bf 258}, 338--358, 2014.
  \end{itemize}

  \underline{Presentations}
  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods for Sparse Linear Systems
    (M.L. Pasini, M. Benzi, T. Evans, S. Hamilton, S. Slattery), \textit{SIAM
      ALA}, October 30, 2015.
  \item Parallel Algorithms for the Monte Carlo Synthetic Acceleration Linear
    Solver Method (S. Slattery, T. Evans, S. Hamilton), \textit{SIAM CSE},
    Salt Lake City, 2015.
  \item Iterative Performance of Monte Carlo Linear Solver Methods
    (M.L. Pasini), \textit{SIAM CSE}, Salt Lake City, 2015.
  \item A multilevel Monte Carlo method for linear systems (S. Slattery,
    T. Evans, S. Hamilton), \textit{Copper Mountain Conference on Iterative
      Solvers}, Copper MT, CO, 2014.
  \item Monte Carlo Synthetic Acceleration as approximate polynomial
    preconditioning (S. Hamilton, T. Evans, S. Slattery), \textit{Copper
      Mountain Conference on Iterative Solvers}, Copper MT, CO, 2014.
  \item Monte Carlo Linear Solvers (S. Slattery, T. Evans, S. Hamilton),
    \textit{NC State Seminar}, Nov. 4, 2014.
  \item Monte Carlo Linear Solvers (S. Hamilton, T. Evans, S. Slattery),
    \textit{Emory Mathematics Dept. Seminar}, Sept. 19, 2014.
  \item Monte Carlo Methods for Linear Systems (T. Evans), \textit{MIT Nuclear
      Engineering Dept Seminar Series}, Sept. 30, 2014.
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\section{Transition and Roadmap}
%%---------------------------------------------------------------------------%%
\subsection{ECP Application Requirements}
%%---------------------------------------------------------------------------%%

\begin{frame}{Connection to ECP Requirements}

  \item Inter-Operability
    \begin{itemize}
    \item MCLS built using Trilinos solver API
    \item Utilizes Hypre/ParaSAILS for AINV preconditioning
    \item Enables testing of MC solver algorithms in a wide variety of
      applications, ie. anyone who uses Trilinos
    \end{itemize}

  \item Scalability
    \begin{itemize}
    \item Multilevel inter-node domain decomposition methods show excellent
      scalability
    \item GPU-enabled on-node MC kernel yields measures speedups of 30 -
      50$\times$
    \end{itemize}

  \item Resiliency
    \begin{itemize}
    \item Replicated subsets enable useful work to occur at the same time as
      providing redundancy
    \item Yields protection against hard errors
    \end{itemize}
  \end{itemize}

\begin{frame}{Inter-Operatibility Example}

\begin{frame}{ECP Coordinated Reasearch}

  \underline{\bf Goal}: Extend the general algorithmic MC exascale methods
  w.r.t. scalability and resiliency into a broader application space

  MC methods are used in many DOE application areas:
  \begin{itemize}
  \item high-energy density physics
  \item astrophysics
  \item nuclear engineering
  \item particle accelerators and neutron scattering
  \item quantum chemistry
  \item visualization
  \item fluid dynamics
  \item computational biology
  \end{itemize}

  Much of the algorithmic research performed in MCREX is directly applicable
  to the use of MC methods in these other application areas

\begin{frame}{Continued Research for MC Linear Solvers}

  Investigate use in time-dependent applications, e.g. non-linear radiation
  hydrodynamics for high-energy density physics applications:
  \begin{itemize}
  \item Used as inner linear solver in a non-linear (Newton) iteration
  \item Time-dependent applications generally have strong diagonal-dominance
    that makes MCSA competitive with Krylov methods
  \item MCSA has resiliency and parallel efficiency benefits over subspace
    methods
  \end{itemize}

\begin{frame}{Building on ASCR Funding}

  MCREX has leveraged and built on work derived from several ASCR projects:
  \begin{itemize}
  \item Trilinos
  \item XPress
  \item xSim
  \item Ideas
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\section*{Extra Slides}
%%---------------------------------------------------------------------------%%
